Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network

### 2013-05-24 doc/paper.tex
finished addressing most of tysons comments and added (almost) finalized sky plots and CI tables

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/VIRGO Network}
\author{Carl L. Rodriguez \altaffilmark{1}
%\email{rodriguez@u.northwestern.edu}
,Benjamin Farr \altaffilmark{1}
%\email{bfarr@u.northwestern.edu} 
Vivien Raymond \altaffilmark{1,2}
Will M. Farr \altaffilmark{1}
%\email{w-farr@northwestern.edu} 
Tyson Littenberg\altaffilmark{1}
%\email{tyson.littenberg@northwestern.edu}
Diego Fazi\altaffilmark{1}
%\email{d-fazi@northwestern.edu}
Vicky Kalogera\altaffilmark{1}}
%\email{vicky@northwestern.edu}

@@

Within the next 5 years, the Advanced LIGO/VIRGO network will have reached a sensitivity sufficient to enable the routine detection of gravitational waves.  Beyond the initial detection, the scientific promise of these instruments relies on the efficiency of the parameter estimation techniques being currently developed.  The majority of this effort has been towards the detection and characterization of gravitational waves from compact binary coalescence, i.e. the coalescence of black holes and/or neutron stars.  While several previous studies have focused on the parameter estimation abilities of advanced detectors, the vast majority have relied on approximation techniques (such as the Fisher Matrix) when characterizing the uncertainties.  Here, we report the statistical uncertainties that will be achievable, in practice, using the parameter estimation code \texttt{LALInference} to recover the parameters of binary neutron star systems detected in various configurations of the Advanced LIGO/VIRGO network.  We find that the recovery of the individual masses will be, on average, sufficient to differentiate neutron stars from black holes by the masses alone.  We also report the average uncertainties on the sky-locations, luminosity distance, and orbital inclinations of randomly generated sources that can be recovered by the expected three and four-detector networks.
\end{abstract}

@@

 By 2016, the field of gravitational-wave astrophysics will
 mature into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and VIRGO detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense work is being undertaken
 to prepare to characterize and extract as much physical 
 information as possible from these signals.

The mergers of binary neutron stars are expected to be the most ubiquitous
source in the advanced detector era \carl{[CITE]}.   Models from stellar 
evolution suggest that the number of binary neutron star mergers within 
Advanced LIGO/VIRGO's detection horizon could reach into the hundreds per 
year \carl{[CITE]}.  Although the peak sensitivity of ground-based detectors
is not focused on the frequency at which BNS systems merge, it should still
be possible to extract information about both strong field gravitational 
physics \carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation 
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the observations of multiple BNS
systems will provide key insight into the evolution of binary systems in the field \carl{[CITE]},
and in globular clusters \carl{[CITE]}. As such, BNS systems will likely 
form the ``bread and butter'' of the compact object efforts in the coming 
years.

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream with a bank of theoretical templates, the roughly 16KHz
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection candidate is found.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to nearly identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
and proper sampling, we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systems.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/VIRGO era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have pointed out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which investigates
the BNS parameter estimation capabilities of Advanced LIGO/VIRGO via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations of certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

In this paper, we perform a systematic study of the statistical uncertainties with
which the Advanced LIGO/VIRGO network will be able to measure the basic parameters of 
binary neutron stars.  In Section \ref{PEsection}, we describe
the machinery of our parameter estimation code, \texttt{LALInference}, and the associated
MCMC sampler, \texttt{lalinference\_mcmc}, as well as the frequency-domain gravitational-wave template we employ.   
In Section \ref{resultsSection}, we qualitatively analyze the posterior probability density functions for BNS systems with different masses and extrinsic parameters.  The results are divided into three parameter sets of interest: the recovery of the mass parameters (Section \ref{massSection}), the recovery of the orbital inclination and luminosity distance (Section \ref{idSection}), and the isolation of sources in the sky (Section \ref{skySection}).   Finally we provide quantitative 1-dimensional confidence intervals on the parameter recovery in Section \ref{ciSection}.  We assume that $G=c=1$ throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with mean zero.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as proportional to the probability
that our signal is Gaussian once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for our template
waveforms.  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Rule to obtain the posterior probability of the system
parameters $\thpara$ given the signal $s$ as
\begin{align}
  p(\thpara | s) &= \frac{p(\thpara)p(s | \thpara)}{p(s)} \nonumber\\
  & \propto p(\thpara) \exp\left[-\frac{1}{2}\big < s - h(\thpara) | s-h(\thpara) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\thpara)$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.  Our prior
information can come from physics limits in the parameter space, or
from our \textit{a priori} knowledge of astrophysical systems. If we
pick a set of parameters $\thpara$ such that
$h(\thpara) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\thpara_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

The LIGO Algorithm Library parameter estimation code, \texttt{LALInference}, is designed as 
a unified framework for gravitational-wave parameter estimation.  By using a common setup for 
waveform generation, PSD estimation, data handling, and other associated techniques from gravitational-wave parameter estimation, \texttt{LALInference} allows the implementation of multiple parameter samplers in using identical parameter setups, including nested sampling (\texttt{lalinference\_nest}) and Markov-Chain Monte Carlo (\texttt{lalinference\_mcmc}).  We elect to use the MCMC sampler for this study.  \texttt{lalinference\_mcmc} is based upon the previously described code, 
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The MCMC employs a Metropolis-Hastings sampling algorithm, which can be described as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\thpara|s\right)$.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/VIRGO network.  
A description of the parameter estimation capabilities, c.f. 2010, can be found
in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = A f^{-7/6}e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $A \propto \chmass^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5} M^{-1/5},
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is empirically superior when proposing MCMC parameter jumps, and is more physically intuitive.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

\noindent where the $\alpha_{k}$ coefficients are taken from the pN expansion to order $n/2$.  See \cite{BuonannoWaveform} for a description and comparison of different waveform families.

The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants
of integration, referred to as the chirp time and coalescence phase,
respectively.  Although uninteresting physically, they must be
accounted for in any parameter estimation study of the waveform phase.

@@

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector as 
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}^{*}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

@@

\begin{equation}
\rho_{\text{network}} = \sqrt{\sum_i \rho_{i}}
\label{SNRnetwork}
\end{equation}

@@

For this study, we consider four seperate populations of BNS systems, with component mass 
combinations of $1M_{\odot}/1M_{\odot}$, $1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, 
and $2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals distributed randomly
in sky location, polarization, inclination, time-of-arrival, and coalescence phase.  The luminosity 
distance, $D$, was selected to provide a network SNR of $\rho_{\text{network}}=20$ for each source.
While unphysical from an astrophysical perspective (binary sources should be 
distributed evenly in comoving volume), this choice allows us to explore the parameter estimation
in the context of viable detection candidates.

Finally, note that the power-spectral density that defines the inner product \eqref{innerProduct} is simply 
the time-averaged sensitivity of a given detector to a specific frequency.  Ignoring non-Gaussian glitches,
 any stretch of data should contain a specific noise realization drawn from a Gaussian colored by the PSD.
   However, what we are interested in is not the parameter estimation that can occur in a specific draw
    from this probability of noises, but the uncertainties averaged over all possible noise realizations.
    \footnote{Similarly to how the Fisher matrix formalism can provide the Cramer-Rao bound averaged over
     all noise realizations} Since our noise model is assumed to be Gaussian, it is unnecessary to recover the
      parameters from a large sample of identical waveforms in different noise; we can simply assume the 
      noise \emph{is} the mean defined by the power-spectral density.  This ``zero-noise'' parameter estimation
       provides a statical statement on the measurable uncertainties, and is what is reported here.  In Section
        \ref{noiseSection} we provide an example of parameter estimation in several non-zero noise realizations
compared to the zero-noise equivelant.

Of the 9 parameters in the domain of the waveform, only 6 are particularly physically
interesting: the masses of the two binaries, $M_1$ and $M_2$, the orbital inclination,
$\iota$, the angular position on the sky, $\alpha$ and $\delta$, and the luminosity 
distance of the source, $D$.  While the coalescence phase $\phi_c$, the coalescence time
$t_c$, and the wave polarization $\psi$ must be included in any parameter estimation of the 
waveform, they do not encode any information of particular astrophysical interest.
\footnote{For systems whose components are non-spinning.  The presence of intrinsic angular 
momentum will, in general, couple to the coalescence phase and wave polarization.}

In Figure~\ref{9dPDF}, we provide an example of the 9, 1-dimensional marginalized posterior 
probability density functions recovered from a single $1.4M_{\odot}/1.4M_{\odot}$ BNS system. 
 These PDFs are typical of what will be produced by parameter estimation studies in the advanced 
 detector era.  Notice that the peak of several parameters, including the chirp mass, $\chmass$, 
 appears to be displaced from the true values in red.  This effect is due to the reduction of the
  9-dimensional PDF to a series of marginalized 1-dimensional PDFs.  For instance, the 1-dimensional 
  PDF for chirp mass is marginalized via

\noindent where the notation $\thpara \setminus \chmass$ implies all
parameter of \eqref{parameterspace} except $\chmass$.  Other parameter and higher-dimensional marginalizations follow a similar convention.

\begin{figure*}[ht!]
  \centering
 \includegraphics[trim=2cm 0cm 2cm 0cm, clip=false,scale=0.7]{newMasses.pdf}
 \caption{Mass PDFs for the four BNS systems, averaged over each of the 40 injections.  The average is reported since, in practice, there was virtually no difference between the recovered PDFs for different systems with identical masses and SNRs \carl{[QUANTIFY]}.  Notice how the PDFs for systems with identical mass ratios but different total masses are identical when normalized to the true total mass.}
  \label{metaMassPDFs}
\end{figure*}

Of the 9 variables in our parameter space \eqref{parameterspace}, the two of immediate physical
interest are the two parameters, $\chmass$ and $q$, or correspondingly the direct masses 
of the two companions, $M_1$ and $M_2$.    As stated above, the ability of
Advanced LIGO/VIRGO to construct a population of BNS masses will be critical to the science
potential of gravitational-wave astronomy.

We find that there is virtually no difference between the mass PDFs of our different injected signals
within each mass bin.  While this may initially seem disconcerting, it is to be expected.  Recall that of the BNS signals were injected with a network SNR of 20 into a zero-noise detector realization.  Furthermore, note that the mass parameters are the only two which directly effect the phase of the TaylorF2 waveform \eqref{phase}.  Therefore, as long as the injected mass parameters are identical within each mass system, and as long as the injected SNR is identical, the amount of recoverable information in each mass parameter should be \emph{almost} identical, and for a sufficiently converged MCMC chain, the recovered posterior PDF should be \emph{almost} identical.

The ``almost'' in the previous sentence is to draw attention to one critical, neglected fact: while the mass parameters are independent of the injected signals position with respect to the detector network, the other parameters are certainly not.  Obviously many of the other parameters (i.e. sky location, polarization, distance, inclination, and time-of-arrival) all depend on the orientation of the binary with respect to each of the individual detectors.  Since our 9-dimensional parameter space can be highly correlated, it is entirely reasonable to assume that cross-correlations between the extrinsic parameters and the mass parameters will cause variation in the recovery of the individual injection PDFs.

In practice, it was found that the cross-correlations between the extrinsic parameters and signal time-of-arrival were several orders of magnitude \carl{[QUANTIFY]} below the uncertainties in $\chmass$ and $q$.  It should be noted that, in practice, this will \emph{not} be the case: in reality, the noise power-spectral density of each detector varies as a function of time, including both natural variations in sensitivity and detector glitches \carl{[CITE]}.  In such a case, the base sensitivity of a gravitational-wave detector to even the easiest to recover parameters will vary as a function of position and time.

In Figures~\ref{1414masses} and~\ref{125masses}, we show the marginalized 2D posterior PDFs of our mass parameters a prototypical equal mass and unequal mass binaries.  We include the PDF in both the $\chmass$-$q$ space (relevant for the waveform and the MCMC algorithm), and the more physically interesting component mass space ($M_1$-$M_2$).  Although only the $1.4M_{\odot}/1.4M_{\odot}$ system is included in Figure \ref{1414masses}, the PDF is virtually identical to the other equal mass cases, modulo a scaling factor.  Again, this is unsurprising due to the identical SNR of each of the injected signals.

>>>>

\begin{document}
\title{Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network}
\author{Carl L. Rodriguez \altaffilmark{1}
%\email{rodriguez@u.northwestern.edu}
,Benjamin Farr \altaffilmark{1}
%\email{bfarr@u.northwestern.edu} 
Vivien Raymond \altaffilmark{1,2}
Will M. Farr \altaffilmark{1}
%\email{w-farr@northwestern.edu} 
Tyson Littenberg\altaffilmark{1}
%\email{tyson.littenberg@northwestern.edu}
Diego Fazi\altaffilmark{1}
%\email{d-fazi@northwestern.edu}
Vicky Kalogera\altaffilmark{1}}
%\email{vicky@northwestern.edu}

@@

Within the next five years, the Advanced LIGO/Virgo network will have reached a sensitivity sufficient to enable the routine
 detection of gravitational waves.  Beyond the initial detection, the scientific promise of these instruments relies the 
 parameter estimation techniques being currently developed.  The majority of this effort has been towards the detection
   and characterization of gravitational waves from compact binary coalescence, i.e. the coalescence of black holes and/or 
   neutron stars.  While several previous studies have focused on the parameter estimation abilities of advanced detectors, 
   the majority have relied on approximation techniques such as the Fisher Matrix.  Here, we report the statistical 
   uncertainties that will be achievable, in practice, using the parameter estimation techniques which will be applied to 
   detections of binary neutron star systems.  We find the recovery of the individual masses to be within fractionally within 
   10\% at the 65\% confidence level for equal-mass systems, and within 2\% for unequal-mass systems.
     We also report the average uncertainties on the sky-locations, luminosity distance, and orbital inclinations of randomly 
     distributed sources that can be achieved by different network configurations. \carl{[PUT IN SA VALUES FOR SKY LOCATION]}
\end{abstract}

@@

 By the end of the decade, the field of gravitational-wave astrophysics will
 mature into an ideal tool for 
 exploring the gravitational side of the universe. Within the next few
 years, the first generation of gravitational-wave detectors capable 
 of regularly resolving astrophysical sources will come online \carl{[CITE]}.
 The Advanced LIGO and Virgo detectors will provide the first insights
 into the final moment of binary compact object mergers, including the 
 merger of binary neutron star systems.  Intense preparations are underway
 to characterize and extract as much physical 
 information as possible from these signals.

The mergers of binary neutron stars are expected to be the most common
source in the advanced detector era \carl{[CITE]}.   Models from stellar 
evolution suggest that the number of binary neutron star mergers within 
Advanced LIGO/Virgo's detection horizon could reach into the hundreds per 
year \carl{[CITE]}.  Although the peak sensitivity of ground-based detectors
is not focused on the frequency at which BNS systems merge, it could still
be possible to extract information about both strong field gravitational 
physics \carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation 
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the observations of multiple BNS
systems will provide key insight into the evolution of binary systems in the field \carl{[CITE]},
and in globular clusters \carl{[CITE]}. As such, BNS systems will likely 
form the ``bread and butter'' of the compact binary coalescence detection effort in the coming 
years.

Of course one must distinguish between the detection of 
such events, and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched filtering approach.  
By comparing the data stream with a bank of theoretical templates, the 
 time-series data can be searched for candidate signals at a sufficiently
  rapid rate to analyze months of data.  When the overlap between one of the template waveforms and the 
detector output is sufficiently large (97\%), a detection candidate is found.  However,
the parameter space of these signals can be highly degenerate, with several 
locations in parameter space corresponding to nearly identical waveforms.  In 
order to fully realize the science potential of these instruments, we must 
perform a full exploration of the parameter space for each detection.  By 
analyzing the parameter space of each candidate with a Bayesian inference technique 
 we will be able to make precise, scientifically meaningful
statements about the physics of these detectible systems.  To 
that end, we employ a Markov-Chain Monte Carlo sampling algorithm, 
\texttt{lalinference\_mcmc}, included in the LIGO Application Library parameter
estimation code, \texttt{LALInference}, to analyze the parameter space of binary neutron star systems.

By employing the full parameter-estimation machinery that will eventually be
used in the Advanced LIGO/Virgo era, our results give the first estimate at the 
realistic capabilities of advanced ground-based detectors to characterize BNS systems.
  Until recently, the majority of studies have employed the Fisher matrix formalism which was 
first adapted for gravitational-wave parameter estimation by \citep{FinnDetection}.
While each of these studies (\cite{PoissonWill,CutlerFlanagan,ArunPE}, among 
many others) have pointed out the limitations 
and flaws of the Fisher Information Matrix, there has been no study which investigates
the BNS parameter estimation capabilities of Advanced LIGO/Virgo via the techniques
that will eventually be employed.  The recent work of \cite{Vallisneri} and our 
own work \citep{Inadequacies} has demonstrated that the Fisher matrix cannot even
be treated as a lower bound on the standard deviations of certain parameters
(particularly the masses) measurable in gravitational-wave detectors.

In this paper, we perform a systematic study of the statistical uncertainties with
which the Advanced LIGO/Virgo network will be able to measure the basic parameters of 
binary neutron stars.  In Section \ref{PEsection}, we describe
the machinery of our parameter estimation code, \texttt{LALInference}, and the associated
MCMC sampler, \texttt{lalinference\_mcmc}, as well as the frequency-domain 
gravitational-wave template we employ.   In Section \ref{resultsSection}, we qualitatively 
analyze the posterior probability density functions for BNS systems with different masses 
and extrinsic parameters.  The results are divided into three parameter sets of interest: the
recovery of the mass parameters (Section \ref{massSection}), the recovery of the orbital
inclination and luminosity distance (Section \ref{idSection}), and the localization of sources 
in the sky (Section \ref{skySection}).   Finally we provide quantitative 1-dimensional confidence
 intervals on the parameter recovery in Section \ref{ciSection}.  These results, contained
 in Tables \ref{ciTableIntrinsic} and \ref{ciTableExtrinsic}, are intended as a reference for the 
 optimal BNS parameter estimation capabilities of Advanced LIGO/Virgo.
   We assume that $G=c=1$ throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation. 
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave
network can be written as a combination of a gravitational
waveform $h_0$ and the noise of the detector $n$.  We further assume
that this noise is stationary and Gaussian with zero mean.  Therefore,
the detector output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since our noise model is Gaussian, we can write the
probability of a specific signal realization $s$ as proportional to the probability
that our residual is Gaussian distributed once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for our template
waveforms, and $p(s | \thpara)$ is the likelihood of the signal $s$ given the parameters $\thpara$.
  The inner product, $\left< ~|~ \right> $, is defined using the noise
spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density as a function
of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the Fourier
transforms of the time-domain signals $a(t)$ and $b(t)$.

Once we have the likelihood of the signal \eqref{likelihood}, we
employ Bayes Theorem to obtain the posterior probability of the system
parameters $\thpara$ given the signal $s$ as
\begin{align}
  p(\thpara | s) &= \frac{p(\thpara)p(s | \thpara)}{p(s)} \nonumber\\
  & \propto p(\thpara) \exp\left[-\frac{1}{2}\big < s - h(\thpara) | s-h(\thpara) \big > \right] ,
  \label{posterior}
\end{align}
where $p(\thpara)$ are the prior probabilities on our
source parameters and $p(s)$ is a normalization constant.

We are interested in the posterior $p(\thpara | s)$ as it encodes information about
prior state of knowledge about the problem.  For instance, since it is assumed that compact binaries
form homogeneously in the local universe, we can safely assume that the population of
sources will be distributed homogeneously in volume, or evenly in $D^2$. 
 Our prior information can come from mathematical limits in the parameter space or
from our \textit{a priori} knowledge of astrophysical systems.  We employ priors which are

\begin{itemize}
\item uniform in component masses from $0.8M_{\odot} \leq M_{1,2} \leq 30M_{\odot}$, with a minimum chirp mass  \eqref{eqRatioCM} of 0.6 $M_{\odot}$,
\item uniform in volume, which implies a luminosity distance prior of $p(D) = D^2 dD$, and
\item uniform in all other parameters.
\end{itemize}

If we pick a set of parameters $\thpara$ such that
$h(\thpara) = h_0$, then the posterior \eqref{posterior}
will be near a global maximum; however, the presence of noise will in
general deflect the maximum of our posterior away from the value at
$\thpara_0$. That is, in the presence of noise, there is
no guarantee that the maximum-likelihood of our signal corresponds to
the true parameters of the system.

@@

The LIGO Algorithm Library parameter estimation code, \texttt{LALInference}, is designed as 
a unified framework for gravitational-wave parameter estimation.  By using a common setup for 
waveform generation, PSD estimation, data handling, and other associated techniques from gravitational-wave parameter estimation, \texttt{LALInference} allows the implementation of multiple parameter samplers in using identical parameter setups, including Nested Sampling (\texttt{lalinference\_nest})\carl{[CITE]} and Markov-Chain Monte Carlo (\texttt{lalinference\_mcmc}).  We elect to use the MCMC sampler for this study.  \texttt{lalinference\_mcmc} is based upon the previously described code, 
\texttt{SpinSpiral} \citep{spinspiral2009, spinspiral2010}.  The MCMC employs a Metropolis-Hastings sampling algorithm, which can be described as follows \citep{Gilks99}:

@@

The above procedure is designed to record a chain of samples whose
distribution is $p\left(\thpara|s\right)$.  By drawing a sufficient ($\sim 1000$) number
 of effective samples from the posterior, the chains traces out the functional form of the posterior,
 gathering more samples from regions with high posterior probability.  Depending on
the proposal distribution, $q$, the convergence (mixing) of the chain
may be rapid or slow.  We employ multiple optimization techniques,
including both specially-crafted $q$ and parallel tempering, to ensure
adequate mixing of the Markov Chains throughout our parameter space.
The details of the algorithm can be found in
\citep{Sluys08,spinspiral2009, spinspiral2010}.  Both samplers were tuned and 
developed during the last science run of the Initial LIGO/Virgo network.  
A description of the parameter estimation capabilities of these two samplers with respect to
real gravitational-wave data, as well as a more detailed description of the algorithms and
checks for convergence, can be found in \cite{S6PE}.

@@

We use a frequency domain waveform accurate up to $3.5$
post-Newtonian (pN) order in phase and amplitude.  We restrict ourselves to
quasi-circular, non-spinning waveforms as a simplifying assumption.  The standard
form of our waveform model, known as the \textit{TaylorF2}
approximant, is calculated via the stationary-phase approximation.
  In this setup, the gravitational-wave
amplitude is given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase reference point, which to lowest order
takes the form $A \propto f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, $D$ is
the luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the
\textit{mass ratio} and \textit{chirp mass}, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5} M^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 / M^2$.  We elect to use $q$ as it is more physically intuitive, and the prior $p(q)$ is uniform when the component
mass priors are uniform.}  The stationary phase then becomes an expansion in the 
Newtonian orbital velocity, $v=(\pi M f)^{1/3}$, 
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} + \frac{3}{128}\left(\frac{M}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

\noindent where the $\alpha_{k}$ coefficients are taken from the pN expansion to order $n/2$.  See \cite{BuonannoWaveform} for a description and comparison of different waveform families.  The terms $t_c$ and $\phi_0$ in equation (\ref{phase}) are constants of integration, referred to as the chirp time and coalescence phase, respectively.

@@

We define the signal-to-noise ratio (SNR) of a gravitational wave in a
single detector as 
\begin{equation}
  \rho \equiv \frac{4}{\sigma} \int^{\infty}_{0}\frac{| \tilde{s}(f)\tilde{h}(f)|}{S_{n}(f)}df
  \label{formalSNR}
\end{equation}
where $\rho$ is the SNR and $\tilde s(f)$ and $\tilde{h}^{*}(f)$ are
the frequency-domain signal and template, respectively, and the normalization $\sigma$ is
given by
\begin{equation}
  \sigma^2 = 4\int^{\infty}_{0}\frac{| \tilde{h}(f)|^2}{S_n(f)}df.
  \label{SNRnorm}
\end{equation}

@@

\begin{equation}
\rho_{\text{network}} = \sqrt{\sum_i \rho_{i}^2}
\label{SNRnetwork}
\end{equation}

@@

For this study, we consider four seperate populations of BNS systems, with component mass 
combinations of $1M_{\odot}/1M_{\odot}$, $1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, 
and $2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals distributed randomly
in sky location, polarization, inclination, time-of-arrival, and coalescence phase.  The luminosity 
distance, $D$, was selected to provide a network SNR of $\rho_{\text{network}}=20$ for each source.
While unphysical from an astrophysical perspective (binary sources should be 
distributed evenly in comoving volume), this choice allows us to explore the parameter estimation
in the context of optimal detection candidates.

Finally, note that the power-spectral density that defines the inner product \eqref{innerProduct} is simply 
the time-averaged sensitivity of a given detector to a specific frequency.  Ignoring non-Gaussian glitches,
 any stretch of data should contain a specific noise realization drawn from a Gaussian colored by the PSD.
   However, what we are interested in is not the parameter estimation that can occur in a specific draw
    from this probability of noises, but the uncertainties averaged over all possible noise
     realizations.\footnote{Similarly to how the Fisher matrix formalism can provide the Cramer-Rao bound averaged over
     all noise realizations} Since our noise model is assumed to be Gaussian, it is unnecessary to recover the
      parameters from a large sample of identical waveforms in different noise; we can simply assume the 
      noise \emph{is} the mean defined by the power-spectral density.  This ``zero-noise'' parameter estimation
       provides a statistical statement on the measurable uncertainties, and is what is reported here.  In Section
        \ref{noiseSection} we provide an example of parameter estimation in several non-zero noise realizations
compared to the zero-noise equivelant.

\begin{comment}
\subsection{Testing MCMC Convergence}

Each set of MCMC samples is tested for convergence by examining both the Gelman-Rubin diagnostic and the auto-correlation length of the MCMC chains, the details of which can be found in \carl{[CITE]} and \carl{[CITE]}, respectively.  However, there is an additional check we can consider: the standard matched filtering searches employed by the LIGO/Virgo collaboration assume that a detection will occur when one of the template waveforms matches the detector data with a signal overlap of 97\% or better \carl{[CITE]}.  That is, for a template waveform $h^{t}(f)$ and detector output $s(f)$, an overlap of
